---
title: "MIS_Pronos_Suavizamiento_media_movil_1"
author: "Sergibar"
date: "2023-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Suavizamiento mediante media movil 

Comenzaremos instalando las liberias necesarias para el desarrollo 


```{r}

library(tidyverse)      # data manipulation and visualization
library(lubridate)      # easily work with dates and times
library(fpp2)           # working with time series data
library(zoo) 
library(ggplot2)
library(dplyr)


library(astsa, quietly=TRUE, warn.conflicts=FALSE)
library(knitr)
library(printr)
library(plyr)
library(gridExtra)
library(reshape2)
library(TTR)


```


Vamos a importar el documento csv que contiene la demanda de gas natural en el sector eléctrico Mexicano 

```{r}
Demanda_electrico <-read.csv("C:\\Users\\sergi\\OneDrive\\Documentos\\MIS_UNAM\\Segundo_semestre\\Pronosticos_UNAM\\Demanda_electrico_2022_full2.csv", header= TRUE)
```

Comprobemos que el documento Demanda_electrico se haya importado correctamente

```{r}
head(Demanda_electrico)
summary(Demanda_electrico)
```
Tengamos una "primer vista rápida" de los datos del documento Demanda_electrico
```{r}

Demanda_electrico<- ts(Demanda_electrico, frequency = 12, start =c(2005,1))
Demanda_electrico
summary(Demanda_electrico)
plot(Demanda_electrico)

```


Para "descomponer" la serie podemos: 
Source: https://rpubs.com/davoodastaraky/TSA1


Decomposing Seasonal Data:
A seasonal time series, in addition to the trend and random components, also has a seasonal component. Decomposing a seasonal time series means separating the time series into these three components. In R we can use the decompose() function to estimate the three components of the time series

```{r}
Demanda_electricoDC <- decompose(Demanda_electrico)
Demanda_electricoDC
```


Now let´s plot the (trend, seasonal and random components of the decompose analysisi)

```{r}
plot(Demanda_electricoDC)
```


###Seasonally Adjusting
If you have a seasonal time series, you can seasonally adjust the series by estimating the seasonal component, and subtracting it from the original time series. We can see below that time time series simply consists of the trend and random components.



Calculemos los valores de la data, restando "el componente de seaonalidad" 
```{r}
Demanda_electrico_Season <- Demanda_electrico - Demanda_electricoDC$seasonal
Demanda_electrico_Season
plot.ts(Demanda_electrico_Season, main = "Demanda sin componente de seasonalidad", col="blue")
```



¿Cómo se comparan las series con el efecto y sin el efecto de la estacionalidad? 
```{r}
plot(Demanda_electrico, type = "l", col = "red", xlab = "DATE", ylab = "Demand", main = "Demanda electrico con y sin seasonalidad")
lines(Demanda_electrico_Season, col = "blue")
legend("bottomright", legend = c("Demanda electrico", "Demanda electrico sin seasonalidad"), col = c("red", "blue"), lty = 1)
```
 
 
 
Calculemos los valores de la data, restando "el componente de trend" 
```{r}
Demanda_electrico_Trend <- Demanda_electrico - Demanda_electricoDC$trend
Demanda_electrico_Trend
plot.ts(Demanda_electrico_Trend, main = "Demanda sin componente de trend", col="pink")
```


 
Calculemos los valores de la data, restando al que ya se le hab[ia restado el season, ahora tambien el "el componente de trend" 
```{r}
Demanda_electrico_Sea_and_Trend <- Demanda_electrico_Season - Demanda_electricoDC$trend
Demanda_electrico_Sea_and_Trend
plot.ts(Demanda_electrico_Sea_and_Trend, main = "Demanda sin componente season ni componente trend", col="violet")
```
```{r}
plot(Demanda_electrico_Sea_and_Trend, type = "l", col = "violet", xlab = "DATE", ylab = "Demand", main = "Demanda electrico y demanda sin seanonalidad ni trend")
lines(Demanda_electrico, col = "red" , ylim(-1000,1000))
legend("bottomleft", legend = c("Demanda electrico sin seasonalidad ni trend","Demanda electrico 'original' " ), col = c("violet","red"), lty = 1)
``


```{r}

plot(Demanda_electrico, type = "l", col = "red", xlab = "DATE", ylab = "Demand", main = "Demanda electrico con y sin seasonalidad $ sin seanonalidad ni trend")
lines(Demanda_electrico_Season, col = "blue")
lines(Demanda_electrico_Sea_and_Trend, col = "violet")
legend("bottomleft", legend = c("Demanda electrico", "Demanda electrico sin seasonalidad","Demanda electrico sin seasonalidad ni trend" ), col = c("red", "blue","violet"), lty = 1)


# Set the y-axis limits for the third data set
par(new = TRUE)
plot(NA, ylim = c(-1000, 1000), axes = FALSE, xlab = "", ylab = "")
lines(Demanda_electrico_Sea_and_Trend, col = "violet")

```


Decomposing non-Seasonal Data
Recall that non-seasonal time series consist of a trend component and a random component. Decomposing the time series involves tying to separate the time series into these individual components.

One way to do this is using some smoothing method, such as a simple moving average. The SMA() function in the TTR R package can be used to smooth time series data using a moving average. The SMA function takes a span argument as n order. To calculate the moving average of order 5, we set n = 5.

Source: https://rpubs.com/davoodastaraky/TSA1



Calculating and ploting the SMA function (con lag n=3)

```{r}
Demanda_electrico_MA_3 <- SMA(Demanda_electrico_Season, order=3)
Demanda_electrico_MA_3
plot(Demanda_electrico_MA_3)
```



Calculating and ploting the SMA function (con lag n=5)

```{r}
Demanda_electrico_MA_5 <- SMA(Demanda_electrico_Season, order=5)
Demanda_electrico_MA_5
plot(Demanda_electrico_MA_5)

```


```{r}
plot(Demanda_electrico, type = "l", col = "red", xlab = "DATE", ylab = "Demand", main = "Demanda electrico'original' vs MA smooth ")
lines(Demanda_electrico_MA_3, col = "green")
lines(Demanda_electrico_MA_5, col = "brown")
legend("bottomright", legend = c("Demanda electrico", "Demanda smooth MA(3)", "Demanda smooth MA(5)"), col = c("red", "green", "brown"), lty = 1)
```


### wEIGHTED MOVING AVERAGE

Calculating and ploting the WMA function (con lag n=3) & wts = c(1, 2, 3))


```{r}
Demanda_electrico_WA_3 <- WMA(Demanda_electrico_Season, order=3)
Demanda_electrico_WA_3
plot(Demanda_electrico_WA_3)
```




####COMPARANDO MA_3 vs WA_3
```{r}
plot(Demanda_electrico_MA_3, type = "l", col = "green", xlab = "DATE", ylab = "Demand", main = "Demanda electrico' smooth MA_3 vs WA_3 ")
lines(Demanda_electrico_WA_3, col = "gold")
lines(Demanda_electrico, col = "red")
legend("bottomright", legend = c("Demanda smooth MA(3)","Demanda smooth WA(3)", "Demanda electrico"), col = c("green", "gold", "red"), lty = 1)
```



### EXPONENTIAL MOVING AVERAGE

Calculating and ploting the EMA function (con lag n=3) & ratio = 0.2)

```{r}
Demanda_electrico_EA_0.2 <- EMA(Demanda_electrico_Season, order=3, ratio =0.2 )
Demanda_electrico_EA_0.2
plot(Demanda_electrico_EA_0.2)
```


####COMPARANDO TODO

```{r}
plot(Demanda_electrico_MA_3, type = "l", col = "green", xlab = "DATE", ylab = "Demand", main = "Demanda electrico' smooth MA_3 vs WA_3 ")
lines(Demanda_electrico_WA_3, col = "gold")
lines(Demanda_electrico, col = "red")
lines(Demanda_electrico_EA_0.2, col = "darkblue")
legend("bottomright", legend = c("Demanda smooth MA(3)","Demanda smooth WA(3)", "Demanda electrico", "Demanda electrico EMA"), col = c("green", "gold", "red", "darkblue"), lty = 1)
```

