---
title: "Pronosticos_Class_21_Marzo"
author: "Sergi"
date: "2023-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




## APLICATION OF SMOTHING METHODS IN R  

```{r}

library(tidyverse)      # data manipulation and visualization
library(lubridate)      # easily work with dates and times
library(fpp2)           # working with time series data
library(zoo)
library(ggplot2)
library(plyr)
library(dplyr)
library(knitr)
library(TTR)
library(hydroGOF)


```


### Let´s start wiht Demanda_electrico

We first import our data 
```{r}
#URL HPi5
#Demanda_electrico_importado<-read.csv("C:\\Users\\llell\\Documents\\MIS\\Second_semester\\Pronosticos_UNAM_HPi5\\pronosticos_UNAM_git\\pronosticos_UNAM_gitHub\\Demanda_electrico_2022_full1.csv", header= TRUE)

#URL HPi3
Demanda_electrico_importado<-read.csv("C:\\Users\\sergi\\OneDrive\\Documentos\\MIS_UNAM\\Segundo_semestre\\Pronosticos_UNAM_HPi3\\pronosticos_UNAM_git\\Demanda_electrico_2022_full1.csv", header= TRUE)
```


We checked the importing has worked 
```{r}
head(Demanda_electrico_importado)
summary(Demanda_electrico_importado)
typeof(Demanda_electrico_importado)
dim(Demanda_electrico_importado)
```



We add a new columns in a data type of data by transforming our 'original' Date columns into an as.date column 

```{r}
Demanda_electrico_importado$as.date <- as.Date(Demanda_electrico_importado$Date, format = "%m/%d/%Y")
head(Demanda_electrico_importado)
tail(Demanda_electrico_importado)
typeof(Demanda_electrico_importado)
dim(Demanda_electrico_importado)

```


#### Let´s see How it looks like time serie 

```{r}
Demanda_electrico_importado.ts<- ts(Demanda_electrico_importado$Demanded_Gas, frequency = 12, start =c(2005,1), end =c(2022,8))
head(Demanda_electrico_importado.ts)

typeof(Demanda_electrico_importado.ts)
dim(Demanda_electrico_importado.ts)
```


Then we plot the 'original demanded gas in electric time series'
```{r}
plot(Demanda_electrico_importado.ts, col = "red", main = "Demanda electrico 'original' ")

```


Para "descomponer" la serie podemos: 
Source: https://rpubs.com/davoodastaraky/TSA1


Decomposing Seasonal Data:
A seasonal time series, in addition to the trend and random components, also has a seasonal component. Decomposing a seasonal time series means separating the time series into these three components. In R we can use the decompose() function to estimate the three components of the time series

```{r}
Demanda_electricoDC <- decompose(Demanda_electrico_importado.ts)
Demanda_electricoDC
```


Now let´s plot the (trend, seasonal and random components of the decompose analysisi)

```{r}
plot(Demanda_electricoDC)
```


###Seasonally Adjusting
If you have a seasonal time series, you can seasonally adjust the series by estimating the seasonal component, and subtracting it from the original time series. We can see below that time time series simply consists of the trend and random components.



Calculemos los valores de la data, restando "el componente de seaonalidad" 
```{r}
Demanda_electrico_SINSeason <- Demanda_electrico_importado.ts - Demanda_electricoDC$seasonal
Demanda_electrico_SINSeason
plot.ts(Demanda_electrico_SINSeason, main = "Demanda sin componente de seasonalidad", col="blue")
```

|

¿Cómo se comparan las series con el efecto y sin el efecto de la estacionalidad? 
```{r}
plot(Demanda_electrico_importado.ts, type = "l", col = "red", xlab = "DATE", ylab = "Demand", main = "Demanda electrico con y sin seasonalidad")
lines(Demanda_electrico_SINSeason, col = "blue")
legend("bottomright", legend = c("Demanda electrico", "Demanda electrico sin seasonalidad"), col = c("red", "blue"), lty = 1)
```
 
 
 
Calculemos los valores de la data, restando "el componente de trend" 
```{r}
Demanda_electrico_SINTrend <- Demanda_electrico_importado.ts - Demanda_electricoDC$trend
Demanda_electrico_SINTrend
plot.ts(Demanda_electrico_SINTrend, main = "Demanda sin componente de trend", col="pink")
```


 
Calculemos los valores de la data, restando al que ya se le habia restado el season, ahora tambien el "el componente de trend" 
```{r}
Demanda_electrico_SIN_Sea_NI_Trend <- Demanda_electrico_SINSeason - Demanda_electricoDC$trend
Demanda_electrico_SIN_Sea_NI_Trend
plot.ts(Demanda_electrico_SIN_Sea_NI_Trend, main = "Demanda sin componente season ni componente trend", col="violet")
```


```{r}

plot(Demanda_electrico_importado.ts, type = "l", col = "red", xlab = "DATE", ylab = "Demand", main = "Demanda electrico con y sin seasonalidad $ sin seanonalidad ni trend")
lines(Demanda_electrico_SINSeason, col = "blue")
lines(Demanda_electrico_SIN_Sea_NI_Trend, col = "violet")
legend("bottomright", legend = c("Demanda electrico", "Demanda electrico sin seasonalidad","Demanda electrico sin seasonalidad ni trend" ), col = c("red", "blue","violet"), lty = 1)


# Set the y-axis limits for the third data set
par(new = TRUE)
plot(NA, ylim = c(-1000, 1000), axes = FALSE, xlab = "", ylab = "")
lines(Demanda_electrico_Sea_and_Trend, col = "violet")

```





### Replication of  Holt Winter´s method in R

Then we are able to apply the HoltWinters method 

```{r}
Demanda_electrico_HW <- HoltWinters(Demanda_electrico_importado.ts)
head(Demanda_electrico_HW)
tail(Demanda_electrico_HW)
typeof(Demanda_electrico_HW)
```


Let´s see only the SMOTHED VALUES
```{r}
Smoothed_values_Demanda_electrico_HW <- Demanda_electrico_HW$x
Smoothed_values_Demanda_electrico_HW
head(Smoothed_values_Demanda_electrico_HW)
tail(Smoothed_values_Demanda_electrico_HW)
typeof(Smoothed_values_Demanda_electrico_HW)
dim(Smoothed_values_Demanda_electrico_HW)

```


SMOOTHED VALUS ARE EQUAL THAN ORIGINAL ONES
```{r}
plot(Demanda_electrico_importado.ts, col = "red", main = "Demanda electrico 'original' ")
lines(Smoothed_values_Demanda_electrico_HW , col='pink')
```



LEt´s plot the forecast
```{r}
plot(Demanda_electrico_HW)

```

```{r}
Demanda_electrico_HW$fitted
```

##### HAGAMOS COMO TAL EL FORECAST DEL MODELO HOLT-WINTER PARA EL CASO DE DEMANDA_ELECTRICO
```{r}
Demanda_electrico_HW_forecast <- forecast(Demanda_electrico_HW, h=12, level=0.95)
Demanda_electrico_HW_forecast
```

Let´s plot the forecast generated by HoltWinters model in  R 

```{r}
plot(Demanda_electrico_HW_forecast)
```






### Now with Demanda_petrolero

We first import our data 
```{r}
#URL HPi5
#Demanda_electrico_importado<-read.csv("C:\\Users\\llell\\Documents\\MIS\\Second_semester\\Pronosticos_UNAM_HPi5\\pronosticos_UNAM_git\\pronosticos_UNAM_gitHub\\Demanda_electrico_2022_full1.csv", header= TRUE)

#URL HPi3
Demanda_petrolero_importado<-read.csv("C:\\Users\\sergi\\OneDrive\\Documentos\\MIS_UNAM\\Segundo_semestre\\Pronosticos_UNAM_HPi3\\pronosticos_UNAM_git\\Demanda_petrolero_2022_full1.csv", header= TRUE)
```




We checked the importing has worked 
```{r}
head(Demanda_petrolero_importado)
summary(Demanda_petrolero_importado)
typeof(Demanda_petrolero_importado)
dim(Demanda_petrolero_importado)
```


We add a new columns in a data type of data by transforming our 'original' Date columns into an as.date column 

```{r}
Demanda_petrolero_importado$as.date <- as.Date(Demanda_petrolero_importado$Date, format = "%m/%d/%Y")
head(Demanda_petrolero_importado)
tail(Demanda_petrolero_importado)
typeof(Demanda_petrolero_importado)
dim(Demanda_petrolero_importado)

```


#### Let´s see How it looks like time serie 

```{r}
Demanda_petrolero_importado.ts<- ts(Demanda_petrolero_importado$Demanded_Gas, frequency = 12, start =c(2005,1), end =c(2022,8))
Demanda_petrolero_importado.ts

typeof(Demanda_petrolero_importado.ts)
dim(Demanda_petrolero_importado.ts)
```


Then we plot the 'original demanded gas in electric time series'
```{r}
plot(Demanda_petrolero_importado.ts, col = "black", main = "Demanda petrolero 'original' ")

```


Para "descomponer" la serie podemos: 
Source: https://rpubs.com/davoodastaraky/TSA1



REMOVIENDO LOS NAs de Demanda_petrolero_importado.ts
```{r}

Demanda_petrolero_importado.ts <- na.omit(Demanda_petrolero_importado.ts)
Demanda_petrolero_importado.ts
```


```{r}
length(Demanda_petrolero_importado.ts)
```


```{r}
Demanda_petrolero_importado.ts <- as.numeric(Demanda_petrolero_importado.ts)
Demanda_petrolero_importado.ts
```


```{r}
Demanda_petrolero_importado.ts <- na.omit(Demanda_petrolero_importado.ts)
Demanda_petrolero_importado.ts
```



```{r}
Demanda_petroleroDC <- decompose(as.numeric(Demanda_petrolero_importado.ts))
Demanda_petroleroDC
```



```{r}
Demanda_petroleroDC <- decompose(Demanda_petrolero_importado.ts)
Demanda_petroleroDC
```

This will decompose your time series using stl() and plot the decomposed time series. You can replace Demanda_petrolero_importado.ts 

The error message series is not periodic or has less than two periods occurs when the data you are giving either lacks a frequency or the frequency is greater than one-half of the length of the vector

```{r}
# Decompose your time series using stl()
Demanda_petrolero_importado.stl <- stl(Demanda_petrolero_importado.ts, s.window = "periodic")

# Plot the decomposed time series
plot(Demanda_petrolero_importado.stl)
```



### Now with Demanda_residencial


We first import our data 
```{r}
#URL HPi5
#Demanda_electrico_importado<-read.csv("C:\\Users\\llell\\Documents\\MIS\\Second_semester\\Pronosticos_UNAM_HPi5\\pronosticos_UNAM_git\\pronosticos_UNAM_gitHub\\Demanda_electrico_2022_full1.csv", header= TRUE)

#URL HPi3
Demanda_residencial_importado<-read.csv("C:\\Users\\sergi\\OneDrive\\Documentos\\MIS_UNAM\\Segundo_semestre\\Pronosticos_UNAM_HPi3\\pronosticos_UNAM_git\\Demanda_residencial_2022_full1.csv", header= TRUE)
```



We checked the importing has worked 
```{r}
head(Demanda_residencial_importado)
summary(Demanda_residencial_importado)
typeof(Demanda_residencial_importado)
dim(Demanda_residencial_importado)
```


We add a new columns in a data type of data by transforming our 'original' Date columns into an as.date column 

```{r}
Demanda_residencial_importado$as.date <- as.Date(Demanda_residencial_importado$Date, format = "%m/%d/%Y")
head(Demanda_residencial_importado)
tail(Demanda_residencial_importado)
typeof(Demanda_residencial_importado)
dim(Demanda_residencial_importado)

```


#### Let´s see How it looks like time serie 

```{r}
Demanda_residencial_importado.ts<- ts(Demanda_residencial_importado$Demanded_Gas, frequency = 12, start =c(2005,1), end =c(2022,8))
Demanda_residencial_importado.ts

typeof(Demanda_residencial_importado.ts)
dim(Demanda_residencial_importado.ts)
```


Then we plot the 'original demanded gas in electric time series'
```{r}
plot(Demanda_residencial_importado.ts, col = "violet", main = "Demanda residencial 'original' ")

```


Para "descomponer" la serie podemos: 
Source: https://rpubs.com/davoodastaraky/TSA1



REMOVIENDO LOS NAs de Demanda_residencial_importado.ts
```{r}

Demanda_residencial_importado.ts <- na.omit(Demanda_residencial_importado.ts)
Demanda_residencial_importado.ts
```


```{r}
Demanda_residencialDC <- decompose(Demanda_residencial_importado.ts)
Demanda_residencialDC
```


Now let´s plot the (trend, seasonal and random components of the decompose analysisi)

```{r}
plot(Demanda_residencialDC)
```


###Seasonally Adjusting
If you have a seasonal time series, you can seasonally adjust the series by estimating the seasonal component, and subtracting it from the original time series. We can see below that time time series simply consists of the trend and random components.



Calculemos los valores de la data, restando "el componente de seaonalidad" 
```{r}
Demanda_residencial_SINSeason <- Demanda_residencial_importado.ts - Demanda_residencialDC$seasonal
Demanda_residencial_SINSeason
plot.ts(Demanda_residencial_SINSeason, main = "Demanda sin componente de seasonalidad", col="lightblue")
```




