---
title: "Pronosticos_class_14_Marzo_2023_Modelos_Lineales_varias_variables"
author: "Sergibar"
date: "2023-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Continuación de Modelos lineales (Ahora multivariados)

Se había visto que para los modelos lineales se puede hacer transformaciones como Ln(y)

#### Entonces para los modelos lineales con más de una variable 
$$ y = b0 +b1*x1 + b2*x2 + b3*x3 ... bn*xn$$


Vamos a hacer uso de la base de datos mtcars

```{r}
data("mtcars")
head(mtcars)
```

```{r}
summary(mtcars)
```

```{r}
dim(mtcars)
plot(mtcars)
```
Se puede hacer una tabla de correlacion entre los pares de variables

```{r}
cor(mtcars)
```

#### Intento de PCA

```{r}
install.packages("corrr")
library('corrr')
```




```{r}
install.packages("ggcorrplot")
library(ggcorrplot)
```

FactoMineR package in R

Mainly used for multivariate exploratory data analysis; the factoMineR package gives access to the PCA module to perform principal component analysis.

```{r}
install.packages("FactoMineR")
library("FactoMineR")
```


Normalizing the data
As stated early in the article, PCA only works with numerical values. So, we need to get rid of the Country column. Also, the Total column is not relevant to the analysis since it is the linear combination of the remaining numerical variables.

The code below creates new data with only numeric columns.


Now, the normalization can be applied using the scale() function. 


```{r}
mtcars_noMPG <- mtcars[, -1] 
head(mtcars_noMPG)
```


```{r}
mtcars_normalized <- scale(mtcars_noMPG)
head(mtcars_normalized)
```


Compute the correlation matrix
Even though the covariance matrix is stated in the previous five steps, the correlation also can be used, and can be computed using the cor() function from the corrr package. The ggcorrplot() can be applied then for better visualization.   


```{r}
corr_matrix <- cor(mtcars_normalized)
ggcorrplot(mtcars_normalized)
```


The result of the correlation matrix can be interpreted as follow: 

The higher the value, the most positively correlated the two variables are.
The closer the value to -1, the most negatively correlated they are.


Applying PCA
Now, all the resources are available to conduct the PCA analysis. First, the princomp() computes the PCA, and summary() function shows the result.



```{r}
data.pca <- princomp(corr_matrix)
summary(data.pca)
```


Each component explains a percentage of the total variance in the data set. 

It’s great to have the first two components, but what do they really mean? 

This can be answered by exploring how they relate to each column using the loadings of each principal component. 


```{r}
data.pca$loadings[, 1:9]
```

```{r}
install.packages("factoextra")  # Install the package
library(factoextra)            # Load the package


fviz_eig(data.pca, addlabels = TRUE)
```



```{r}
fviz_cos2(data.pca, choice = "var", axes = 1:9)
```




### Continuamos con la clase sobre modelo lineales

```{r}
ajuste1 <- lm(mpg~1++hp+wt+cyl+disp, data=mtcars)
ajuste1
```

```{r}
summary(ajuste1)
```




```{r}

ajuste2 <- lm(mpg~1+wt+am, data=mtcars)
ajuste2
summary(ajuste2)
```




```{r}

ajuste3 <- lm(mpg~1+wt+cyl, data=mtcars)
ajuste3
summary(ajuste3)
```


```{r}
plot(ajuste3)[0]
```



```{r}

ajuste4 <- lm(mpg~1+wt+qsec, data=mtcars)
ajuste4
summary(ajuste4)
```



```{r}

ajuste5 <- lm(mpg~1+wt+qsec+am, data=mtcars)
ajuste5
summary(ajuste5)
```


Ajuste con el resultado de PCA 
```{r}
ajuste_PCA1 <- lm(mpg~1+cyl+disp, data=mtcars)
ajuste_PCA1
summary(ajuste_PCA1)
```


#### Ejemplo con data UScrime

```{r}
library(MASS)
data(UScrime)
head(UScrime)
```

```{r}
summary(UScrime)
dim(UScrime)
```

```{r}
plot(UScrime)
```



```{r}
cor(UScrime)
```

```{r}
help('UScrime')
```



```{r}
ajuste_crime1 <- lm(y~1++M+U1+Time, data=UScrime)
ajuste_crime1
summary(ajuste_crime1)
```




```{r}
UScrime_drop <- UScrime[, -which(names(UScrime) == "y")]  # Drop the "y" variable

head(UScrime_drop)
```


```{r}
UScrime_normalized <- scale(UScrime_drop)
head(UScrime_normalized)
```


Compute the correlation matrix
Even though the covariance matrix is stated in the previous five steps, the correlation also can be used, and can be computed using the cor() function from the corrr package. The ggcorrplot() can be applied then for better visualization.   


```{r}
corr_matrix <- cor(UScrime_normalized)
ggcorrplot(UScrime_normalized)
```


The result of the correlation matrix can be interpreted as follow: 

The higher the value, the most positively correlated the two variables are.
The closer the value to -1, the most negatively correlated they are.


Applying PCA
Now, all the resources are available to conduct the PCA analysis. First, the princomp() computes the PCA, and summary() function shows the result.



```{r}
data.pca <- princomp(corr_matrix)
summary(data.pca)
```


Each component explains a percentage of the total variance in the data set. 

It’s great to have the first two components, but what do they really mean? 

This can be answered by exploring how they relate to each column using the loadings of each principal component. 


```{r}
data.pca$loadings[, 1:15]
```

```{r}
install.packages("factoextra")  # Install the package
library(factoextra)            # Load the package


fviz_eig(data.pca, addlabels = TRUE)
```


Contribución de variables a la DIM1
```{r}
fviz_cos2(data.pca, choice = "var", Top=5)
```


Contribución de variables a la DIM2
```{r}
fviz_cos2(data.pca, choice = "var", Top=5, axes=2)
```




```{r}
fviz_cos2(data.pca, choice = "var", axes = 1:9)
```



```{r}
ajuste_crime2 <- lm(y~1++GDP+Ineq+Ed, data=UScrime)
ajuste_crime2
summary(ajuste_crime2)
```



```{r}
ajuste_crime3 <- lm(y~1++GDP+Ineq+Ed+So, data=UScrime)
ajuste_crime3
summary(ajuste_crime3)
```


```{r}
ajuste_crime4 <- lm(y~1++GDP+Pop+Time, data=UScrime)
ajuste_crime4
summary(ajuste_crime4)
```



```{r}
ajuste_crime5 <- lm(y~1++GDP+Pop+Ineq+Ed, data=UScrime)
ajuste_crime5
summary(ajuste_crime5)
```




```{r}
ajuste_crime6 <- lm(y~1++GDP+Ineq+Ed, data=UScrime)
ajuste_crime6
summary(ajuste_crime6)
plot(ajuste_crime6)
```


ESTA FUE LA MEJOR A LA QUE YO LLEGUÉ CON AYUDA DEL PCA 
```{r}
ajuste_crime7 <- lm(y~1++GDP+Po2+Ineq, data=UScrime)
ajuste_crime7
summary(ajuste_crime7)
plot(ajuste_crime7)
```

EJEMPLO CON UNA SOLA VARIABLE 

```{r}
ajuste_crimeS <- lm(y~1++GDP, data=UScrime)
ajuste_crimeS
summary(ajuste_crimeS)
plot(ajuste_crimeS)
```


EJEMPLO CON UNA SOLA VARIABLE 


```{r}
ajuste_crimeS2 <- lm(y~-1++GDP, data=UScrime)
ajuste_crimeS2
summary(ajuste_crimeS2)
plot(ajuste_crimeS2)
```


Modelo de la compañera Karen
```{r}

ajuste_crime8 <- lm(y~1++Po2+Ineq+Pop, data=UScrime)
ajuste_crime8
summary(ajuste_crime8)
plot(ajuste_crime8)
```

#### Vamos a importar datos que nos dió el profre

```{r}
energy_datos <- read.table('C:\\Users\\sergi\\OneDrive\\Documentos\\MIS_UNAM\\Segundo_semestre\\Pronosticos_UNAM_HPi3\\pronosticos_UNAM_git\\energy.txt')
energy_datos
```


Transformando a serie de tiempo

```{r}
energy_datos_ts <- ts(energy_datos, frequency = 12, start=c(1985,1))
energy_datos_ts
plot(energy_datos_ts)
```


```{r}
par(mfrow=c(3,4))

mes <- c("Ener", "Feb", "Mar", "apr", "May", "Jun", "Jul", "Ago", "Sept", "Oct", "Nov", "Dec")

for(j in 1:12)
{
  
  bp = window(energy_datos_ts, start=c(1985,j), delta=1)
  plot(bp, main= paste("Mes", mes[j]), type="o", ylim=c(180,305))
}
```

