---
title: "Pronosticos_class_2_May_ARMA_to_myy_data_part1"
author: "Sergibar"
date: "2023-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Application of ARMA & ARIMA Models to my data 

### The CF and ACF in the desicion of AR and MA significatn paratemers


#### Empezemos con los datos de demanda_electrico 

We first import our data 
```{r}
#URL HPi5
#Demanda_electrico_importado<-read.csv("C:\\Users\\llell\\Documents\\MIS\\Second_semester\\Pronosticos_UNAM_HPi5\\pronosticos_UNAM_git\\pronosticos_UNAM_gitHub\\Demanda_electrico_2022_full1.csv", header= TRUE)

#URL HPi3
Demanda_electrico_importado<-read.csv("C:\\Users\\sergi\\OneDrive\\Documentos\\MIS_UNAM\\Segundo_semestre\\Pronosticos_UNAM_HPi3\\pronosticos_UNAM_git\\Demanda_electrico_2022_full1.csv", header= TRUE)
```


We checked the importing has worked 
```{r}
head(Demanda_electrico_importado)
summary(Demanda_electrico_importado)
typeof(Demanda_electrico_importado)
dim(Demanda_electrico_importado)
```


We add a new columns in a data type of data by transforming our 'original' Date columns into an as.date column 

```{r}
Demanda_electrico_importado$as.date <- as.Date(Demanda_electrico_importado$Date, format = "%m/%d/%Y")
head(Demanda_electrico_importado)
tail(Demanda_electrico_importado)
typeof(Demanda_electrico_importado)
dim(Demanda_electrico_importado)

```


Let´s see How it looks like time serie 

```{r}
Demanda_electrico_importado.ts<- ts(Demanda_electrico_importado$Demanded_Gas, frequency = 12, start =c(2005,1), end =c(2022,8))
head(Demanda_electrico_importado.ts)

typeof(Demanda_electrico_importado.ts)
dim(Demanda_electrico_importado.ts)

Demanda_electrico_importado.ts
```


Then we plot the 'original demanded gas in electric time series'
```{r}
plot(Demanda_electrico_importado.ts, col = "red", main = "Demanda electrico 'original' ")

```


#### Claculo de ACF & PACF para el caso de la data de demanda del sector eléctrico 

```{r}
#  ACF (Autocorrelation function)
acf_demanda_electrico_tot <- acf(na.omit(Demanda_electrico_importado), lag.max = 100)
acf_demanda_electrico_tot

#  PACF (Partial Autocorrelation function)
pacf_demanda_electrico_tot <- pacf(na.omit(Demanda_electrico_importado))
pacf_demanda_electrico_tot
```



Note: In R, the pacf() function does not have a lag.max argument like the acf() function. Instead, the pacf() function computes the partial autocorrelation coefficients up to the maximum lag of the data by default. Therefore, we do not need to specify a lag.max parameter when calling the pacf() function.

The lag.max parameter in the acf() function is used to specify the maximum lag at which to compute the autocorrelation coefficients. For example, setting lag.max = 10 would compute the autocorrelation coefficients up to a lag of 10. This parameter is optional and if not specified, it defaults to min(10, n-1) where n is the length of the input data.

It is worth noting that in practice, the choice of lag.max depends on the characteristics of the time series data and the purpose of the analysis. For instance, if the time series data is known to have a seasonal pattern, we might want to set lag.max to a multiple of the seasonal frequency.





#### Claculo de ACF & PACF para el caso de la data de demanda del sector eléctrico 

```{r}
#  ACF (Autocorrelation function)


acf_demanda_electrico <- acf(Demanda_electrico_importado.ts, lag.max = 100)
acf_demanda_electrico 

#  PACF (Partial Autocorrelation function)
pacf_demanda_electrico <- pacf(Demanda_electrico_importado.ts, lag.max = 100)
pacf_demanda_electrico


```


```{r}
#  ACF (Autocorrelation function)


acf_demanda_electrico <- acf(Demanda_electrico_importado.ts, lag.max = 100, ci.type = "ma")
acf_demanda_electrico 
```

```{r}

#  PACF (Partial Autocorrelation function)
pacf_demanda_electrico <- pacf(Demanda_electrico_importado.ts, lag.max = 100, ci.type = "yvm")
pacf_demanda_electrico

```





